---
title: "PithAnalysis"
author: "Allegra DePasquale"
date: "3/3/2021"
output: html_document
---
```{r}
###library statements
library(rsq)
library(ggplot2)
library(tidyverse)
library(viridis)
library(ggpubr)
library(RColorBrewer)

```

###MODELLING & PLOTTING PITH PATCH VISITS AS A FUNCTION OF FRUIT BIOMASS
```{r}
FPV <- read.csv("PithByMonth_WithZeroes.csv")
biomassData <- read.csv("FruitBiomass.csv")
#FPV_no0 <- read.csv("PithByMonth.csv")

#Renaming columns
FPV <- FPV %>% 
           rename(Year = YearOfFPV,)

FPV <- FPV %>% 
           rename(Month = MonthOfFPV,)

biomassData <- biomassData %>% 
           rename(Year = ï..Year,)

#Summarizing data by month and year          
FPV <- FPV %>% group_by(Month, Year) %>%
  summarise(Count = sum(CountOfTaxon))

#Joining FPV data with fruit biomass data
FPV <- FPV %>% left_join(., biomassData, by = c( "Year" = "Year",
                                                          "Month" = "Month"))

#Setting as factor
FPV$Year <- as.factor(FPV$Year)
FPV$Month <- as.factor(FPV$Month)

#Storing colors by year
cols <- c("2018"= "#3e8672", "2019"= "#ffa93a", "2020"="#af3f23", "2021"="#7caaf0")

#plotting yearly pith patch visits by month + fruit biomass
pith_patch_biomass <- ggplot(FPV, aes(x=Biomass, y=Count)) + 
  geom_point(aes(fill=Year), pch=21, size =3,colour="black") +
  geom_smooth(method=lm) +
  scale_fill_manual(values = cols) +
  labs(y = "Number of monthly pith patch visits", x = "Monthly Fruit Biomass (kg/ha)") +
  theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

ggsave("pith_patch_biomass_plot.pdf", plot = pith_patch_biomass)


#Getting regression summary
FPVmodel <- lm(formula = Count ~ Biomass, data = FPV)
summary(FPVmodel)

#Plotting yearly pith patch visits by month + fruit biomass

#Making sure variables are specified correctly
FPV$Month <- as.factor(FPV$Month)
FPV$Year <- as.factor(FPV$Year)
biomassData$Month <- as.factor(biomassData$Month)
biomassData$Year <- as.factor(biomassData$Year)

#Excluding 2021 bc incomplete
FPV <- FPV[FPV$Year != "2021",]
biomassData <- biomassData[biomassData$Year != "2021",]

#Plotting yearly pith patch visits by month + fruit biomass
pith_by_month_biomass <- FPV %>% ggplot(aes(x = Month, group=1)) +
geom_bar(aes(y=Count*2, col = "Count", fill = "Count"), color = "black", fill = "mediumseagreen", stat = "identity") +
geom_line(aes(y=Biomass, col = "Biomass"), color = "mediumslateblue", size =1)  +
facet_grid(rows = vars(Year), switch = "y") +
scale_y_continuous(sec.axis = sec_axis(~./2, name= "Fruit Biomass"))+
theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), legend.position ="none") 

#Saving plot
ggsave(file = "pith_by_month_biomass.pdf", plot = pith_by_month_biomass)

```


###RELATIVE PITH ANALYSIS###


```{r}
#Reading in data
relative <- read.csv("relative_pith.csv")

#Re-leveling factor for plotting
relative$Species <- factor(relative$Species , levels=c("Ateleia herbert-smithii", "Glircidia sepium", "Lonchocarpus miniflorus", "Luehea speciosa", "Rhedera trinervis", "Stemmadenia obovata", "Tabebuia ochracea", "Bursera simaruba", "Cochlospermum vitifolium", "Mabea occidentalis", "Spondias mombin", "Trichilia americana"))

#Plotting relative pith
relative_pith_plot <- ggplot(relative, aes(x=Species, y = Average.Relative.Pith, fill = DietCategory)) + 
    geom_boxplot() + 
   scale_fill_brewer(palette="Dark2") +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), axis.title.x=element_blank(), axis.text.x = element_text(angle = 45, hjust =1), plot.margin = margin(10, 10, 10, 100), legend.position = "bottom")


#Saving plot
ggsave(file="relative_pith_plot.pdf", plot = relative_pith_plot)

#Comparing means 
compare_means(Average.Relative.Pith ~ DietCategory, data = relative)

#Wilcoxon test
wilcox.test(Average.Relative.Pith ~ DietCategory, data = relative)
```

###MECHANICAL ANALYSIS###

```{r}

#Reading in data
mechanical <- read.csv("pith_mechanical.csv")

#Renaming columns
mechanical <- mechanical %>% 
           rename(
           Species = Species.Code,)

mechanical <- mechanical %>% 
           rename(
           Hardness = Indentation.Hardness..GPa.,)


#Setting as factor
mechanical$Species <- as.factor(mechanical$Species)

#Ordering levels correctly

mechanical$Species <- factor(mechanical$Species , levels=c("AHER", "GSEP", "LMIN", "LSPE", "RTRI", "SOBO", "TOCH", "BSIM", "CVIT", "MOCC", "SMOM", "TAME"))

#Changing level names to be full species names

levels(mechanical$Species) <- c("Ateleia herbert-smithii", "Glircidia sepium", "Lonchocarpus miniflorus", "Luehea speciosa", "Rehdera trinervis", "Stemmadenia obovata", "Tabebuia ochracea", "Bursera simaruba", "Cochlospermum vitifolium", "Mabea occidentalis", "Spondias mombin", "Trichilia americana")

#Plotting hardness boxplots
mechanical_plot <- ggplot(mechanical, aes(x=Species, y=Hardness, fill=DietCategory)) + 
    geom_boxplot() + 
    scale_fill_brewer(palette="Dark2") +
    labs(y = "Hardness (GPa)") +
    facet_grid(Surface ~.) +
    theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), axis.title.x=element_blank(), axis.text.x = element_text(angle = 45, hjust =1), plot.margin = margin(10, 10, 10, 100), legend.position = "bottom")

ggsave(file="mechanical_plot.pdf", plot = mechanical_plot)

#Comparing means (Wilcox test, same results as below)
compare_means(Hardness ~ DietCategory, data = mechanical, 
              group.by = "Surface")


#Wilcoxon tests for inner and outer branch hardness
mech_outer <- mechanical[mechanical$Surface != "Inner",]
mech_inner <- mechanical[mechanical$Surface != "Outer",]

wilcox.test(Hardness ~ DietCategory, data = mech_outer)
wilcox.test(Hardness ~ DietCategory, data = mech_inner)

```

###NUTRITIONAL ANALYSIS###

```{r}
df_nutrition <- read.csv("PithNutrition.csv")


df_nutrition1 <- gather(df_nutrition, key = "Nutrient", value = "Percent",
       CP, AP, NDF, ADF, Lignin, WSC, Fat, Ash)


df_nutrition1$Nutrient <- as.factor(df_nutrition1$Nutrient)
df_nutrition1$DietCategory <- as.factor(df_nutrition1$DietCategory)

df_nutrition1$Nutrient <- factor(df_nutrition1$Nutrient, levels=c("ADF", "AP", "Ash", "CP", "Fat", "Lignin", "NDF", "WSC"))

levels(df_nutrition1$Nutrient) <- c("Acid detergent fiber", "Available protein", "Ash", "Crude protein", "Crude fat", "Lignin", "Neutral detergent fiber", "Water-soluble carbohydrates")


bxp <- ggboxplot(df_nutrition1, x = "Nutrient", y = "Percent", fill = "DietCategory") + 
  scale_fill_brewer(palette="Dark2") +
   theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), axis.title.x=element_blank(), axis.text.x = element_text(angle = 45, hjust =1), plot.margin = margin(10, 10, 10, 100), legend.position = "bottom")

ggsave(file = "pith_nutrition_plot.pdf", plot = bxp)


compare_means(Percent ~ DietCategory, data = df_nutrition1, 
              group.by = "Nutrient")

wilcox.test(Percent ~ DietCategory, data = df_nutrition1)



###CHEMICAL ANALYSIS###

### Datasets
# 122 good peaks
#odor<-read.csv("PithOdor.csv", header=T, na.strings="NA", sep=",")
odor[,c(1:6)]<- lapply(odor[,c(1:6)], factor)
head(odor)

# Calculate log(x+1)-transformed relative peak areas (RPA)
odor2<- data.frame(odor %>%
                    group_by(Sample) %>%
                    summarise(Area=Area,
                              SumArea=sum(Area),
                              logRPA=round(log(((Area/sum(Area))*100)+1), digits=3)) %>%
                    add_tally(name="Npeak") )
head(odor2)

odor<- odor %>% arrange(Sample, Area)
odor2<- odor2 %>% arrange(Sample, Area)
odorRPA<- cbind(odor, odor2[,-(1:2)])
head(odorRPA)

# Sample list including number of peaks
odor.spl<-unique(odorRPA[,c(1:3, 11)])
odor.spl<-arrange(odor.cla, Sample)
head(odor.spl)

hist(odor.spl$Npeak)
shapiro.test(odor.spl$Npeak)    # normal distribution: good


### Testing the effect of diet on sample odor richness
wilcox.test(Npeak~Diet, odor.spl)

# Boxplot (The final version of figure 3 is plotted elsewhere)
odor_plot <- 
  ggplot(odor.spl, aes(x=Species, y=Npeak, fill=Diet)) + 
  geom_boxplot() + 
  scale_fill_brewer(palette="Dark2") +
  ggtitle("Odor richness")  +
  labs(y = "Number of VOCs") +
  theme_classic() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), 
                     axis.title.x=element_blank(), axis.text.x = element_text(angle = 45, hjust =1), legend.position = "bottom")
plot(odor_plot)


### Testing the effect of diet on sample odor composition 
# Calculate Bray-Curtis dissimilarity index
library(vegan)

odor.mat<- xtabs(logRPA~Sample+Peak, odorRPA)
odor.vdim<-vegdist(odor.mat, method="bray")

# Analysis of similarity (ANOSIM)
anosim(odor.vdim, odor.cla2$Diet, permutations=999, distance="bray") #R=0.44, P=0.001


### VOC composition
# VOC list dataset
odor.class<-unique(odor[,c(1,2,3,5,6)])
head(odor.class)

odor.class2<- data.frame(odor.class %>%
                     group_by(Sample, Species, Diet, Class) %>%
                       summarize(Count = n()))
head(odor.class2)

# Mean + SE VOCs
odor.classM<- odor.class2 %>% group_by(Diet, Class) %>%  summarise(mean=mean(Count), se=round(sd(Count)/sqrt(sum(n())), digits=3))
odor.classM


# Barplot chemical classes (MARCH 2023 edit)
head(odor.class2)
levels(odor.class2$Class)

Class2<- odor.class2$Class
levels(Class2)<- c("Aliphatics", "Benzenoids\nand\nPhenylpropanoids", "Miscellaneous\ncyclic\ncompounds", 
                   "Nitrogen-\ncontaining\ncompounds", "Terpenoids", "Unidentified\ncompounds")
odor.class2<- mutate(odor.class2, Class2=Class2)
head(odor.class2)

CompClass<-
ggplot(odor.class2, aes(x=Diet, y=Count, fill=Diet)) + 
  geom_boxplot() + 
  facet_grid(.~Class2)+
  scale_fill_brewer(palette="Dark2") +
  labs(y = "Number of compounds") +
  theme_classic() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), legend.title=element_blank(),
                     panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), legend.position = "bottom",
                     axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())
plot(CompClass)


### Suppl Figure 1: Terpenoids in pith samples
terp<-read.csv("PithOdor_Terpenoids.csv", header=T, na.strings="NA", sep=",") 
terp[,c(1:6)]<- lapply(terp[,c(1:6)], factor)
head(terp)

terp2<- data.frame(terp %>% group_by(Diet, Species, Identity) %>% summarize(Count = n()))
terp2<- mutate(terp2, Percent=round(Count*100/sum(Count), digits=1))
head(terp2)

# Reorder species names by non-dietary then dietary
Tlevels<-c("A.her","G.sep","S.obo","B.sim","C.vit","S.mom","T.ame")

CompTerp<-
ggplot(terp2, aes(y=Count, x=factor(Species, Tlevels), fill=Diet))+
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette="Dark2") +
  guides(fill='none')+
  labs(y="Number of compounds") +
  facet_wrap(Identity~., ncol=4)+
  theme_classic() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
                     axis.title.x=element_blank(), axis.text.x = element_text(angle = 45, hjust =1))
plot(CompTerp)





















